---
title: "TE DEA"
output: html_notebook
author: Raquel Garza
---

## Some functions for visualization
```{r class.source = 'fold-hide'}
## getSignName ##
# Get significantly different gene names. 
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# x = results object from deseq
# p = padj threshold for significance
# l = log2FC threshold for significance
getSignName <- function(x,p,l=0) {
  up <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange > l,]
  down <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange < -l,]
  return(list(up=rownames(up),down=rownames(down)))
}
## getAverage ##
# Get average expression (normalized by median of ratios) of each of the conditions in a deseq object.
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# dds = deseq object
getAverage <- function(dds) {
  baseMeanPerLvl <- sapply( levels(dds$condition), function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$condition == lvl] ) )
  baseSDPerLvl <- sapply( levels(dds$condition), function(lvl) apply( counts(dds,normalized=TRUE)[,dds$condition == lvl],1,sd ) )
  colnames(baseSDPerLvl) <- paste("st.dev:",colnames(baseSDPerLvl),sep="")
  return(list(Mean=baseMeanPerLvl,SD=baseSDPerLvl))
}

meanPlot_cus <- function(exp,test,c1 = "condition 1",c2 = "condition 2",p=.05,l=0,id=F, ttl="", 
                         repel=TRUE, col1="firebrick3", col2="steelblue4", col3="black", highlights=NA){
  sign <- getSignName(x = test,p = p,l = l)
  u <- sign$up
  d <- sign$down
  
  #color up and down sign..
  colVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = col1,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = col2, no =col3))
  colVec[is.na(colVec)] <- "steelblue" ## if NA make sure it's not counted as <p
  #size of points
  cexVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = 0.35,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = 0.35, no = 0.3))
  
  exp_log <- as.data.frame(log2(exp[,c(c1, c2)]))
  exp_log$Name <- rownames(exp_log)
  
  exp_log$reg <- factor(ifelse(exp_log$Name %in% u, paste('upregulated in ', c1, ' (', length(u), ')', sep =''),
                               ifelse(exp_log$Name %in% d, paste('downregulated in ', c1,' (', length(d), ')', sep =''), paste('not significant', ' (', (nrow(test) - length(u) - length(d)), ')', sep=''))))
  
  library(ggrepel)
  if(repel == TRUE){
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), label=Name, color=reg)) + geom_point(aes(size=cexVec))+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")+ geom_text_repel(data = subset(exp_log, Name %in% u | Name %in% d),direction    = "y", nudge_y = 0.4, nudge_x = -0.5)
  }
  else{
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), color=reg)) + geom_point(aes(size=cexVec))+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")
  }
  plt <- plt + labs(x=paste("log2(mean ",c2,")",sep=""), 
                    y=paste("log2(mean ",c1,")",sep=""),
                    title=paste(ttl, paste(c1," vs. ",c2,sep=""), sep = ': '),
                    subtitle=paste("p-adj < ",p,", log2(fc) > ",l,sep=""))+theme(
                      plot.title = element_text( size=14, face="bold"),  panel.grid.major = element_line(colour="white"),
                      panel.grid.minor = element_line(colour="white"), panel.background = element_rect(fill = "white",
                                                                                                       colour = "white",
                                                                                                       size = 0.5, linetype = "solid"),
                      axis.line = element_line(size = 0.5, linetype = "solid",
                                               colour = "black"), 
                      legend.title=element_blank()) 
  
  
  if(id==T) {
    
    identify(log2(exp[,1]),log2(exp[,2]),labels = rownames(exp))
    
  }
  
  if(!is.na(highlights)){
    plt <- plt + geom_point(data=exp_log[highlights,], aes(x=get(c2), y=get(c1)), colour="springgreen3", size=5, shape=1, stroke=2)
  }
  return(plt)
  
}
```

## Reading data and metadata
```{r}
library(data.table)
library(DESeq2)
library(stringr)
library(xlsx)
library(pheatmap)
library(ggpubr)
library(UpSetR)
library(ggplot2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(tidyr)
gene_counts <- fread("/Volumes/MyPassport/xdp/gene_counts/default/gene_count_matrix_2.csv", data.table = F)
colnames(gene_counts)[7:ncol(gene_counts)] <- sapply(str_split(colnames(gene_counts)[7:ncol(gene_counts)], "/"), `[[`, 8)
rownames(gene_counts) <- gene_counts$Geneid

# There was one sample mislabelled...
colnames(gene_counts)[which(colnames(gene_counts) == "JOA5_ZF91g1D_S26")] <- "mislabelled"
colnames(gene_counts)[which(colnames(gene_counts) == "JOA5_ZF91lacZD_S30")] <- "JOA5_ZF91g1D_S26"
colnames(gene_counts)[which(colnames(gene_counts) == "mislabelled")] <- "JOA5_ZF91lacZD_S30"

coldata <- read.xlsx("/Volumes/MyPassport/xdp/metadata.xlsx", sheetIndex = 1)
rownames(coldata) <- coldata$sample
coldata$cell_type_steps <- paste(coldata$cell_type_at_KD, coldata$cell_type, sep="_")
coldata$gene_cell_type_steps <- paste(coldata$gene, coldata$cell_type_steps, sep="_")
coldata$gene_cell_type_steps_dx <- paste(coldata$gene_cell_type_steps, coldata$dx, sep="_")

te_counts <- fread("/Volumes/MyPassport/xdp/TEcounts/unique/TE_count_matrix_2.csv", data.table = F)
colnames(te_counts)[7:ncol(te_counts)] <- sapply(str_split(colnames(te_counts)[7:ncol(te_counts)], "/"), `[[`, 8)
rownames(te_counts) <- te_counts$Geneid

```

## Gene DEA pooling diagnosis

For assessing the changes in the different experiments, correcting for the individuals' diagnosis and the samples' batch if needed.

Outputs DESeq2 objects. This is the same DEA as performed in `gene_DEA_experiments.Rmd` which we will use to normalize by sizeFactors.
```{r}
coldata_experiments <- split(coldata, f = coldata$gene_cell_type_steps)
# iPSC --> NES didnt have a control, so we are using the LZ from the same individual
coldata_experiments$ZNF91_iPSC_NES <- rbind(coldata_experiments$ZNF91_iPSC_NES,
                                            coldata_experiments$ZNF91_NES_NES[which(coldata_experiments$ZNF91_NES_NES$individual == "C1" & coldata_experiments$ZNF91_NES_NES$condition == "Control"),])
for(experiment in names(coldata_experiments)) rownames(coldata_experiments[[experiment]]) <- coldata_experiments[[experiment]]$sample

# Some experiments were sequenced in several sequencing runs. So we need to correct for batch effects in those..
coldata_experiments_batch <- coldata_experiments[c("ZNF91_NES_NES", "TRIM28_NES_NES")]
same_dx <- names(coldata_experiments)[which(names(coldata_experiments) %in% c("ZNF91_iPSC_iPSC", "ZNF91_iPSC_NES", "ZNF91_fbNPC_fbNPC", "ZNF91_Neurons_Neurons"))]
coldata_experiments_no_dx <- coldata_experiments[same_dx]
same_batch <- names(coldata_experiments)[which(!names(coldata_experiments) %in% c("ZNF91_NES_NES", "TRIM28_NES_NES", same_dx))]
coldata_experiments_no_batch <- coldata_experiments[same_batch]

experiment_dds <- list()
for(experiment in names(coldata_experiments_no_batch)) experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(gene_counts[,rownames(coldata_experiments_no_batch[[experiment]])], coldata_experiments_no_batch[[experiment]], design =  ~ dx + condition)
# Run DEA for experiments with possible batch effect
for(experiment in names(coldata_experiments_batch)) experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(gene_counts[,rownames(coldata_experiments_batch[[experiment]])], coldata_experiments_batch[[experiment]], design =  ~ dx + sequencing_run + condition)
# Run DEA for experiments with same dx (we cant correct by sequencing run in the ZNF91 ipsc nes experiment. and the other two experiments in this category were sequenced in the same batch) 
for(experiment in names(coldata_experiments_no_dx)) experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(gene_counts[,rownames(coldata_experiments_no_dx[[experiment]])], coldata_experiments_no_dx[[experiment]], design =  ~ condition)

for(experiment in names(coldata_experiments)) experiment_dds[[experiment]]$condition <- relevel(experiment_dds[[experiment]]$condition, "Control")
for(experiment in names(coldata_experiments)) experiment_dds[[experiment]] <- DESeq(experiment_dds[[experiment]])
experiment_vst <- list()
for(experiment in names(coldata_experiments)) experiment_vst[[experiment]] <- varianceStabilizingTransformation(experiment_dds[[experiment]])
experiment_norm_counts <- list()
for(experiment in names(coldata_experiments)) experiment_norm_counts[[experiment]] <- counts(experiment_dds[[experiment]], normalized = T)
experiment_res <- list()
for(experiment in names(coldata_experiments)) experiment_res[[experiment]] <- lfcShrink(experiment_dds[[experiment]], "condition_KD_vs_Control")
```

## TE DEA 

For assessing the changes in the different experiments, correcting for the individuals' diagnosis and the samples' batch if needed.

Outputs DESeq2 objects and mean plots for >1kbp SVAs.
```{r}
te_experiment_dds <- list()
te_experiment_res <- list()
for(experiment in names(coldata_experiments_no_batch)) te_experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(te_counts[,rownames(coldata_experiments_no_batch[[experiment]])], coldata_experiments_no_batch[[experiment]], design =  ~ dx + condition)
# Run DEA for experiments with possible batch effect
for(experiment in names(coldata_experiments_batch)) te_experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(te_counts[,rownames(coldata_experiments_batch[[experiment]])], coldata_experiments_batch[[experiment]], design =  ~ dx + sequencing_run + condition)
# Run DEA for experiments with same dx (we cant correct by sequencing run in the ZNF91 ipsc nes experiment. and the other two experiments in this category were sequenced in the same batch) 
for(experiment in names(coldata_experiments_no_dx)) te_experiment_dds[[experiment]] <- DESeqDataSetFromMatrix(te_counts[,rownames(coldata_experiments_no_dx[[experiment]])], coldata_experiments_no_dx[[experiment]], design =  ~ condition)
for(experiment in names(coldata_experiments)) te_experiment_dds[[experiment]]$condition <- relevel(te_experiment_dds[[experiment]]$condition, "Control")
for(experiment in names(coldata_experiments)) te_experiment_dds[[experiment]] <- DESeq(te_experiment_dds[[experiment]])

# Subset objects just for SVAs... Otherwise the objects are huge and this becomes extra slow
te_experiment_dds_svas <- list()
te_experiment_exp_svas <- list()
for(experiment in names(coldata_experiments)) te_experiment_dds_svas[[experiment]] <- te_experiment_dds[[experiment]][which(startsWith(rownames(te_experiment_dds[[experiment]]), "SVA")),]
for(experiment in names(coldata_experiments)) te_experiment_exp_svas[[experiment]] <- getAverage(te_experiment_dds_svas[[experiment]])
for(experiment in names(coldata_experiments)) te_experiment_res[[experiment]] <- lfcShrink(te_experiment_dds[[experiment]], "condition_KD_vs_Control")

# Filter for only >1kb SVAs
plus1kb_SVAs <- te_counts[which(startsWith(te_counts$Geneid, "SVA") & te_counts$Length >= 1000), "Geneid"]
plus1kb_SVAs_experiment_res <- list()
for(experiment in names(coldata_experiments)) plus1kb_SVAs_experiment_res[[experiment]] <- te_experiment_res[[experiment]][which(rownames(te_experiment_res[[experiment]]) %in% plus1kb_SVAs),]

te_experiment_exp_plus1kb_sign_svas <- list()
for(experiment in names(coldata_experiments)) te_experiment_exp_plus1kb_sign_svas[[experiment]] <- te_experiment_exp_svas[[experiment]]$Mean[rownames(plus1kb_SVAs_experiment_res[[experiment]]),,drop=F]

# Mean plots for experiments
# pdf("/Volumes/MyPassport/xdp/manuscript/figures/plots/mean_plots_1kb_SVAs.pdf", width = 20, height = 14)
ggarrange(meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$DNMT1_NES_NES, test=plus1kb_SVAs_experiment_res$DNMT1_NES_NES, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - DNMT1', col2 = "gray35", col3 = "firebrick") ,
meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$DNMT1_ZNF91_NES_NES, test=plus1kb_SVAs_experiment_res$DNMT1_ZNF91_NES_NES, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - DNMT1+ZNF91', col2 = "#29358e", col3 = "gray35") ,
meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$ZNF91_NES_NES, test=plus1kb_SVAs_experiment_res$ZNF91_NES_NES, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - ZNF91', col2 = "gray35", col3 = "#29358e") ,
meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$TRIM28_NES_NES, test=plus1kb_SVAs_experiment_res$TRIM28_NES_NES, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - TRIM28', col2 = "gray35", col3 = "#29358e") ,
meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_iPSC, test=plus1kb_SVAs_experiment_res$ZNF91_iPSC_iPSC, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - ZNF91 (iPSC)', col2 = "#29358e", col3 = "gray35") ,
meanPlot_cus(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_NES, test=plus1kb_SVAs_experiment_res$ZNF91_iPSC_NES, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - ZNF91 (iPSC --> NES)', col2 = "#29358e", col3 = "gray35"),
meanPlot_cus(exp = te_experiment_exp_plus1kb_sign_svas$ZNF91_Neurons_Neurons, test=plus1kb_SVAs_experiment_res$ZNF91_Neurons_Neurons, l=1, repel = F, p=0.05, c1='KD', c2='Control',ttl='TE DEA - ZNF91 (Neurons)', col2 = "gray35") )
# dev.off()

```

### Extract names of upregulated SVAs in any of the experiments 

Outputs heatmaps of SVAs upregulated in any of the experiments
```{r}
# DNMT1 NES NES
# ZNF91 NES NES
# ZNF91 iPSC NES
# ZNF91 iPSC iPSC
# DNMT1+ZNF91 NES NES 
upreg_SVAs <- unique(c(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_NES_NES) %in% rownames(te_experiment_res$DNMT1_NES_NES[which(te_experiment_res$DNMT1_NES_NES$log2FoldChange > 1 & te_experiment_res$DNMT1_NES_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_NES_NES) %in% rownames(te_experiment_res$ZNF91_NES_NES[which(te_experiment_res$ZNF91_NES_NES$log2FoldChange > 1 & te_experiment_res$ZNF91_NES_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_ZNF91_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_ZNF91_NES_NES) %in% rownames(te_experiment_res$DNMT1_ZNF91_NES_NES[which(te_experiment_res$DNMT1_ZNF91_NES_NES$log2FoldChange > 1 & te_experiment_res$DNMT1_ZNF91_NES_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_NES) %in% rownames(te_experiment_res$ZNF91_iPSC_NES[which(te_experiment_res$ZNF91_iPSC_NES$log2FoldChange > 1 & te_experiment_res$ZNF91_iPSC_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_iPSC)[which(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_iPSC) %in% rownames(te_experiment_res$ZNF91_iPSC_iPSC[which(te_experiment_res$ZNF91_iPSC_iPSC$log2FoldChange > 1 & te_experiment_res$ZNF91_iPSC_iPSC$padj < 0.05),]))]))

te_experiment_norm <- list()
for(experiment in names(coldata_experiments)){
  te_experiment_norm[[experiment]] <- te_counts[,rownames(coldata_experiments[[experiment]])]
  te_experiment_norm[[experiment]][] <- mapply('/', te_experiment_norm[[experiment]][,rownames(coldata_experiments[[experiment]])], experiment_dds[[experiment]]$sizeFactor[rownames(coldata_experiments[[experiment]])])
  coldata_experiments[[experiment]]$condition <- factor(coldata_experiments[[experiment]]$condition, levels = c("Control", "KD"))
  coldata_experiments[[experiment]]$sample <- factor(coldata_experiments[[experiment]]$sample, levels = coldata_experiments[[experiment]][order(coldata_experiments[[experiment]]$condition),"sample"])
}

levels(coldata_experiments[[experiment]][order(coldata_experiments[[experiment]]$condition),"sample"])
# te_experiment_norm_df <- do.call("rbind", te_experiment_norm)


# pdf("/Volumes/MyPassport/xdp/results/plots/SVAs_each_experiment.pdf")
upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$DNMT1_ZNF91_NES_NES[upreg_SVAs,]) > 0)]
tmp <- pheatmap(te_experiment_norm$DNMT1_ZNF91_NES_NES[upreg_SVAs_expressed,levels(coldata_experiments$DNMT1_ZNF91_NES_NES$sample)], cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$DNMT1_ZNF91_NES_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_experiments$DNMT1_ZNF91_NES_NES$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "DNMT1 + ZNF91 KD (NES NES)")

upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs,]) > 0)]
tmp <- pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed, levels(coldata_experiments$DNMT1_NES_NES$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_experiments$DNMT1_NES_NES$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "DNMT1 KO (NES NES)")

coldata_DNMT1_NES_NES_X1 <- coldata_experiments$DNMT1_NES_NES[which(coldata_experiments$DNMT1_NES_NES$individual == "X1"),]
coldata_DNMT1_NES_NES_X1$sample <- factor(coldata_DNMT1_NES_NES_X1$sample, levels= coldata_DNMT1_NES_NES_X1[order(coldata_DNMT1_NES_NES_X1$condition),"sample"])
upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs, levels(coldata_DNMT1_NES_NES_X1$sample)]) > 0)]
tmp <- pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed, levels(coldata_DNMT1_NES_NES_X1$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_DNMT1_NES_NES_X1$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "X1 - DNMT1 KO (NES NES)")

coldata_DNMT1_NES_NES_C1 <- coldata_experiments$DNMT1_NES_NES[which(coldata_experiments$DNMT1_NES_NES$individual == "C1"),]
coldata_DNMT1_NES_NES_C1$sample <- factor(coldata_DNMT1_NES_NES_C1$sample, levels= coldata_DNMT1_NES_NES_C1[order(coldata_DNMT1_NES_NES_C1$condition),"sample"])
upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs, levels(coldata_DNMT1_NES_NES_C1$sample)]) > 0)]
tmp <- pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed, levels(coldata_DNMT1_NES_NES_C1$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$DNMT1_NES_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_DNMT1_NES_NES_C1$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "C1 - DNMT1 KO (NES NES)")

upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$ZNF91_NES_NES[upreg_SVAs,]) > 0)]
tmp <- pheatmap(te_experiment_norm$ZNF91_NES_NES[upreg_SVAs_expressed, levels(coldata_experiments$ZNF91_NES_NES$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$ZNF91_NES_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_experiments$ZNF91_NES_NES$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "ZNF91 KD (NES NES)")

upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$ZNF91_iPSC_NES[upreg_SVAs,]) > 0)]
tmp <- pheatmap(te_experiment_norm$ZNF91_iPSC_NES[upreg_SVAs_expressed, levels(coldata_experiments$ZNF91_iPSC_NES$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$ZNF91_iPSC_NES[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_experiments$ZNF91_iPSC_NES$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "ZNF91 KD (iPSC --> NES)")

upreg_SVAs_expressed <- upreg_SVAs[which(rowSums(te_experiment_norm$ZNF91_iPSC_iPSC[upreg_SVAs,]) > 0)]
tmp <- pheatmap(te_experiment_norm$ZNF91_iPSC_iPSC[upreg_SVAs_expressed, levels(coldata_experiments$ZNF91_iPSC_iPSC$sample)], cluster_rows = T, cluster_cols = F, show_rownames = F)
pheatmap(te_experiment_norm$ZNF91_iPSC_iPSC[upreg_SVAs_expressed[tmp$tree_row$order], levels(coldata_experiments$ZNF91_iPSC_iPSC$sample)], scale = "row", cluster_rows = F, cluster_cols = F, show_rownames = F,
         main = "ZNF91 KD (iPSC iPSC)")
# dev.off()
```


## All experiments together

Normalize using sizeFactors, and visualize upregulated SVAs (in any experiment, >1kb)
```{r}
# Run DEA for experiments with possible batch effect
gene_dds <- DESeqDataSetFromMatrix(gene_counts[,rownames(coldata)], coldata, design =  ~ dx + sequencing_run + condition)
gene_dds$condition <- relevel(gene_dds$condition, "Control")
gene_dds <- DESeq(gene_dds)

te_norm <- te_counts[,rownames(coldata)]
te_norm[] <- mapply('/', te_norm[,rownames(coldata)], gene_dds$sizeFactor[rownames(coldata)])


coldata$gene_cell_type_steps <- factor(coldata$gene_cell_type_steps, levels = c("ZNF91_Neurons_Neurons",
                                                                                "TRIM28_NES_NES",
                                                                                "ZNF91_NES_NES",
                                                                                "DNMT1_NES_NES",
                                                                                "ZNF91_fbNPC_fbNPC",
                                                                                "ZNF91_iPSC_iPSC",
                                                                                "ZNF91_iPSC_NES",
                                                                                "DNMT1_ZNF91_NES_NES"))

coldata$sample <- factor(coldata$sample, levels= coldata[order(coldata$condition, coldata$gene_cell_type_steps_dx),"sample"])

upreg_SVAs_nes <- unique(c(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_NES_NES) %in% rownames(te_experiment_res$DNMT1_NES_NES[which(te_experiment_res$DNMT1_NES_NES$log2FoldChange > 1 & te_experiment_res$DNMT1_NES_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_NES_NES) %in% rownames(te_experiment_res$ZNF91_NES_NES[which(te_experiment_res$ZNF91_NES_NES$log2FoldChange > 1 & te_experiment_res$ZNF91_NES_NES$padj < 0.05),]))],
                       rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_ZNF91_NES_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$DNMT1_ZNF91_NES_NES) %in% rownames(te_experiment_res$DNMT1_ZNF91_NES_NES[which(te_experiment_res$DNMT1_ZNF91_NES_NES$log2FoldChange > 1 & te_experiment_res$DNMT1_ZNF91_NES_NES$padj < 0.05),]))]))

coldata_nes <- coldata[which(as.character(coldata$gene_cell_type_steps) %in% c("ZNF91_NES_NES", "DNMT1_NES_NES", "DNMT1_ZNF91_NES_NES")),]
coldata_nes <- coldata_nes[c(which(coldata_nes$condition == "Control" & coldata_nes$individual == "C1")[1:4], 
                             which(coldata_nes$gene_cell_type_steps == "ZNF91_NES_NES" & coldata_nes$condition == "KD" & coldata_nes$individual == "C1")[1:4],
                             which(coldata_nes$gene_cell_type_steps == "DNMT1_NES_NES" & coldata_nes$condition == "KD" & coldata_nes$individual == "C1")[1:4],
                             which(coldata_nes$gene_cell_type_steps == "DNMT1_ZNF91_NES_NES" & coldata_nes$condition == "KD" & coldata_nes$individual == "C1")[1:4]),]
coldata_nes$sample <- factor(coldata_nes$sample, levels = as.character(coldata_nes[order(coldata_nes$gene_cell_type_steps, coldata_nes$condition),"sample"]))
pdf("/Volumes/MyPassport/xdp/results/plots/0_SVAs_upreg_NES_ZNF91_DNMT1_DNMT1ZNF91_NES_C1.pdf")
pheatmap(log2(te_norm[upreg_SVAs_nes, levels(coldata_nes$sample)]+0.5), scale = "row", cluster_rows = T, cluster_cols = F, annotation_col = coldata[which(as.character(coldata$gene_cell_type_steps) %in% c("ZNF91_NES_NES", "DNMT1_NES_NES", "DNMT1_ZNF91_NES_NES")),c("gene_cell_type_steps", "condition")],
         show_rownames = F, main = "Upregulated SVAs", show_colnames = F, gaps_col = c(4,4,4, 8,12), color=RColorBrewer::brewer.pal(10, "YlGnBu"))
dev.off()


upreg_SVAs_ips_nes <- unique(c(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_NES)[which(rownames(te_experiment_exp_plus1kb_sign_svas$ZNF91_iPSC_NES) %in% rownames(te_experiment_res$ZNF91_iPSC_NES[which(te_experiment_res$ZNF91_iPSC_NES$log2FoldChange > 1 & te_experiment_res$ZNF91_iPSC_NES$padj < 0.05),]))]))

coldata_ipsc_nes <- coldata[which(as.character(coldata$gene_cell_type_steps) %in% c("ZNF91_iPSC_NES", "ZNF91_iPSC_iPSC", "ZNF91_NES_NES", "DNMT1_ZNF91_NES_NES")),]
coldata_ipsc_nes <- coldata_ipsc_nes[c(which(coldata_ipsc_nes$condition == "Control" & coldata_ipsc_nes$individual == "C1" & coldata_ipsc_nes$gene_cell_type_steps == "ZNF91_iPSC_iPSC")[1:4], 
                                       which(coldata_ipsc_nes$condition == "Control" & coldata_ipsc_nes$individual == "C1" & coldata_ipsc_nes$gene_cell_type_steps == "ZNF91_NES_NES")[1:4],
                                      which(coldata_ipsc_nes$gene_cell_type_steps == "ZNF91_NES_NES" & coldata_ipsc_nes$individual == "C1"  & coldata_ipsc_nes$condition == "KD")[1:4],
                                      which(coldata_ipsc_nes$gene_cell_type_steps == "ZNF91_iPSC_NES" & coldata_ipsc_nes$individual == "C1"  & coldata_ipsc_nes$condition == "KD"),
                                      which(coldata_ipsc_nes$gene_cell_type_steps == "ZNF91_iPSC_iPSC" & coldata_ipsc_nes$individual == "C1"  & coldata_ipsc_nes$condition == "KD"),
                                      which(coldata_ipsc_nes$gene_cell_type_steps == "DNMT1_ZNF91_NES_NES" & coldata_ipsc_nes$individual == "C1"  & coldata_ipsc_nes$condition == "KD")),]

coldata_ipsc_nes$sample <- factor(coldata_ipsc_nes$sample, levels = as.character(coldata_ipsc_nes[order(coldata_ipsc_nes$cell_type_at_KD, coldata_ipsc_nes$condition, coldata_ipsc_nes$gene_cell_type_steps),"sample"]))
pdf("/Volumes/MyPassport/xdp/results/plots/0_SVAs_upreg_iPSC_NES_ZNF91_DNMT1ZNF91_NES_C1.pdf")
pheatmap(log2(te_norm[upreg_SVAs_ips_nes, levels(coldata_ipsc_nes$sample)]+0.5), scale = "row", cluster_rows = T, cluster_cols = F, annotation_col = coldata_ipsc_nes[,c("gene_cell_type_steps", "condition")], show_rownames = F, main = "Upregulated SVAs", show_colnames = F, gaps_col = c(4,4,8,12,12,12,12,12,16,16,20), color=RColorBrewer::brewer.pal(10, "YlGnBu"))
dev.off()

```

log2FC of those SVAs
```{r}
log2FC_upreg_SVAs_ipsc_nes <- data.frame()
log2FC_upreg_SVAs_nes <- data.frame()
for (experiment in names(te_experiment_res)){
  log2FC_upreg_SVAs_ipsc_nes <- rbind(log2FC_upreg_SVAs_ipsc_nes, data.frame(experiment = experiment,
             log2FC = te_experiment_res[[experiment]][upreg_SVAs_ips_nes,"log2FoldChange"]))
  log2FC_upreg_SVAs_nes <- rbind(log2FC_upreg_SVAs_nes, data.frame(experiment = experiment,
             log2FC = te_experiment_res[[experiment]][upreg_SVAs_nes,"log2FoldChange"]))
}
library(ggplot2)


log2FC_upreg_SVAs_ipsc_nes <- log2FC_upreg_SVAs_ipsc_nes[which(log2FC_upreg_SVAs_ipsc_nes$experiment %in% c("ZNF91_iPSC_NES", "ZNF91_iPSC_iPSC", "ZNF91_NES_NES", "DNMT1_ZNF91_NES_NES")),]
log2FC_upreg_SVAs_ipsc_nes$experiment <- factor(log2FC_upreg_SVAs_ipsc_nes$experiment, levels = c("ZNF91_iPSC_iPSC", "ZNF91_iPSC_NES", "ZNF91_NES_NES", "DNMT1_ZNF91_NES_NES"))


log2FC_upreg_SVAs_nes <- log2FC_upreg_SVAs_nes[which(log2FC_upreg_SVAs_nes$experiment %in% c("ZNF91_NES_NES", "DNMT1_NES_NES", "DNMT1_ZNF91_NES_NES")),]
log2FC_upreg_SVAs_nes$experiment <- factor(log2FC_upreg_SVAs_nes$experiment, levels = c("ZNF91_NES_NES", "DNMT1_NES_NES", "DNMT1_ZNF91_NES_NES"))

# pdf("/Volumes/MyPassport/xdp/results/plots/0_SVAs_upreg_iPSC_NES_ZNF91_DNMT1ZNF91_NES_allsamples_boxplot.pdf", height = 3, width = 3)
ggplot(log2FC_upreg_SVAs_ipsc_nes, aes(x=experiment,y=log2FC)) + geom_boxplot() + theme_classic() + labs(x="", y="log2FoldChange") + ggtitle("Upregulated SVAs in iPSC --> NES")
# dev.off()

# pdf("/Volumes/MyPassport/xdp/results/plots/0_SVAs_upreg_NES_ZNF91_DNMT1_DNMT1ZNF91_NES_allsamples_boxplot.pdf", height = 3, width = 3)
ggplot(log2FC_upreg_SVAs_nes, aes(x=experiment,y=log2FC)) + geom_boxplot() + theme_classic() + labs(x="", y="log2FoldChange") + ggtitle("Upregulated SVAs in NES")
# dev.off()
```
